Exploiting Array Syntax in Fortran for Accelerator Programming\footnote*{Foo!}

### 2010-12-19 opencl/papers/foropencl.tex
if someone doesn't know what FORTRAN stands for, they shouldn't be     reviewing papers.
edits

\usepackage{cite}
\usepackage{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{array}
\usepackage{url}
\usepackage{listings}

@@

\title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}

%%
%% reorder appropriately later
%%
\author{\IEEEauthorblockN{Matthew J. Sottile}
\IEEEauthorblockA{Galois, Inc.\\
421 SW 6th Ave. Suite 300 \\
Portland, OR 97204\\
Email: matt@galois.com}
\and
\IEEEauthorblockN{Craig E Rasmussen}
\IEEEauthorblockA{Los Alamos National Laboratory\\
CCS-7, MS B287\\
Los Alamos, NM\\
Email: rasmussn@lanl.gov}
}

@@

\begin{abstract}
Emerging architectures for high performance computing often are well
suited to a data parallel programming model.  This paper presents a
simple programming methodology based on existing languages and compiler
tools that allows programmers to take advantage of these systems.
We will work with the array features of Fortran 90 to show how this
infrequently exploited, standardized language feature is easily
transformed to lower level accelerator code.  Our transformations are
based on a mapping from Fortran 90 to C++ code with OpenCL extensions.
\end{abstract}

\input{intro}
\input{programming-model}
\input{shallow-water}
\input{source-to-source}
\input{static-analysis}

\section{Performance}

@@

%% Explain what we have accomplished (copied from proposal)

%% NEW - CER
The sheer complexity of programming for clusters of many or multi-core
processors with tens of millions threads of execution make the simplicity of
the data parallel model attractive.  Furthermore, the increasing complexity of
todays applications (especially when convolved with the increasing complexity
of the hardware) and the need for portability across hardware architectures
make a higher-level and simpler programming model like data parallel
attractive.

%% NEW - CER
The goal of this work has been to exploit source-to-source transformations that
allow programmers to develop and maintain programs at a high-level of
abstraction, without coding to a specific hardware architecture.
Furthermore these transformations allow multiple hardware architectures
to be targeted without changing the high-level source.  It also removes the
necessity for application programmers to understand details of the accelerator
architecture or to know OpenCL.

% Summarize results

\section{Junk}

\cite{chamberlain04zpl, roth97stencils}

>>>>

\usepackage{cite}
\usepackage{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{url}
\usepackage{listings}

@@

%% Changed by CER
%%\title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}

\title{Exploiting Array Syntax in Fortran for Accelerator Programming\footnote*{Foo!}}

%%
%% reorder appropriately later
%%
\author{\IEEEauthorblockN{Matthew J. Sottile}
\IEEEauthorblockA{Galois, Inc.\\
%421 SW 6th Ave. Suite 300 \\
%Portland, OR 97204\\
Email: matt@galois.com}
\and
\IEEEauthorblockN{Craig E Rasmussen,\\
                  Wayne N. Weseloh,\\
                  Robert W. Robey}
\IEEEauthorblockA{Los Alamos National Laboratory\\
%CCS-7, MS B287\\
%Los Alamos, NM\\
Email: rasmussn@lanl.gov}

\and
\IEEEauthorblockN{Daniel Quinlan}
\IEEEauthorblockA{Lawrence Livermore\\ National Laboratory}

\and
\IEEEauthorblockN{Jeffrey Overbey}
\IEEEauthorblockA{Indiana University}

}

@@

\begin{abstract}
Emerging architectures for high performance computing are well suited
to a data-parallel programming model.  This paper presents a simple
programming methodology that allows programmers to take advantage of
these systems.  We use array constructs in Fortran 90 to show how this
infrequently exploited, standardized language feature is easily
transformed to lower-level accelerator code.  These transformations
are based on a simple mapping from Fortran 90 to OpenCL.  We
demonstrate, using a shallow-water code, that an acceleration up to 40
times can be achieved on GPU hardware with these compiler-generated
transformations.
\end{abstract}

\input{intro}
\input{programming-model}
\input{shallow-water}
\input{source-to-source}
%\input{static-analysis}
\input{performance}

@@

The sheer complexity of programming for clusters of many or multi-core
processors with tens of millions threads of execution make the simplicity of
the data parallel model attractive.  Furthermore, the increasing complexity of
todays applications (especially when convolved with the increasing complexity
of the hardware) and the need for portability across hardware architectures
make a higher-level and simpler programming model like data parallel
attractive.

The goal of this work has been to exploit source-to-source transformations that
allow programmers to develop and maintain programs at a high-level of
abstraction, without coding to a specific hardware architecture.
Furthermore these transformations allow multiple hardware architectures
to be targeted without changing the high-level source.  It also removes the
necessity for application programmers to understand details of the accelerator
architecture or to know OpenCL.

%%TODO \cite{chamberlain04zpl, roth97stencils}

### 2010-12-19 opencl/papers/intro.tex
edits

At the Los Alamos National Laboratory (LANL), as with many supercomputing
facilities today, users have a wide variety of computer platforms from
which to choose.  The most common platform is made up of clusters of
compute nodes with standard multi-core processors.  An increasingly
common feature is that some nodes also have accelerators that range
from the IBM Cell processor (such as the LANL Roadrunner system) to a
variety of GPUs from NVIDIA and AMD.  Some nodes have hardware with
vector instructions and others do not.  The peak performance of these
accelerated nodes often resides in the hundreds of gigaflops.

@@

However, the performance that these accelerator architectures offer
comes at a cost, as processor architectures are trending toward
multiple cores with instances of integrated accelerator units (with
user managed memory) and less of a reliance on superscalar instruction
level parallelism and hardware managed memory hierarchies (such as
traditional caches).  These changes place a heavy burden on
application programmers as they are forced to adapt to the new
systems.  An especially challenging problem faced by application
programmers is not only how to program to these new architectures
(considering the massive scale of concurrency available), but also how
to design programs that are portable across the changing landscape of
computer architectures with unique memory systems and programming
models.  The fundamental question we address is what programming model
and language constructs are best suited to span this set of new
hardware designs.

@@

Transformations for general Fortran procedures are not attempted.
Furthermore, a mechanism to transform the calling site to
automatically invoke OpenCL kernels is not provided at this time.
While it is possible to accomplish this task within ROSE, it is
considered outside the scope of this paper.

@@

Fortran is the oldest high-level programming language in continuous
use since its introduction, and was developed to facilitate the
translation of math formulae into machine code. Fortran was the first
major language to use a compiler to translate from a high-level
program representation to assembly language. Due to its age, it
carries certain arcane baggage.  However
with the introduction of Fortran 90 (and later revisions to the
standard), Fortran became a truly modern programming language. 
It is now modular and has many
object-oriented features.  Most importantly for this work, it now
includes a type system in which rich first-class array data types and
corresponding syntax are part of the language, something that
languages like C continue to lack.  Furthermore, Fortran should be of
interest to those studying parallel programming because of its
functional and data parallel constructs and because of the coarray
notation introduced in Fortran 2008. Unlike languages like C and C++,
Fortran has become a truly
parallel language with features added to recent language standards.

@@

%% NEW - CER
History has also shown that an investment in rewriting code does not guarantee
success either, as seen in an effort at LANL to modernize a legacy Fortran
code with the the newer C++ POOMA framework.  This massive overhaul effort led
to a code that was both slower and less flexible than the original Fortran
\cite{basili08hpc}.  Similar experiences have occurred in the past, notably during the
devlopment of the Sisal language.

>>>>

%At the Los Alamos National Laboratory (LANL), as with many supercomputing
%facilities today, users have a wide variety of computer platforms from
%which to choose.  The most common platform is made up of clusters of
%compute nodes with standard multi-core processors.  An increasingly
%common feature is that some nodes also have accelerators that range
%from the IBM Cell processor (such as the LANL Roadrunner system) to a
%variety of GPUs from NVIDIA and AMD.  Some nodes have hardware with
%vector instructions and others do not.  The peak performance of these
%accelerated nodes often resides in the hundreds of gigaflops.

@@

The performance that new accelerator architectures offer comes at a
cost, as processor architectures are trending toward multiple cores
with instances of integrated accelerator units (with user managed
memory) and less of a reliance on superscalar instruction level
parallelism and hardware managed memory hierarchies (such as
traditional caches).  These changes place a heavy burden on
application programmers as they are forced to adapt to the new
systems.  An especially challenging problem faced by application
programmers is not only how to program to these new architectures
(considering the massive scale of concurrency available), but also how
to design programs that are portable across the changing landscape of
computer architectures with unique memory systems and programming
models.  Can a programmer easily write one program that can run on a
conventional multicore CPU, graphics processing unit, Cell processor,
and one of many emerging many core architectures?  The fundamental
question we address is what programming model and language constructs
are best suited to span this set of new hardware designs.

@@

Transformations for arbitrary Fortran procedures are not attempted.
Furthermore, a mechanism to transform the calling site to
automatically invoke OpenCL kernels is not provided at this time.
While it is possible to accomplish this task within ROSE, it is
considered outside the scope of this paper.

@@

Fortran is the oldest high-level programming language in continuous
use since its introduction, and was developed to facilitate the
translation of math formulae into machine code. Fortran was the first
major language to use a compiler to translate from a high-level
program representation to assembly language. Due to its age, it
carries certain arcane baggage.  However with the introduction of
Fortran 90 (and later revisions to the standard), Fortran became a
truly modern programming language.  It is now modular and has many
object-oriented features.  Most importantly for this work, it now
includes a type system in which rich first-class array data types and
corresponding syntax are part of the language, something that
languages like C continue to lack\footnote{The recent Intel 12.0 C/C++
  compiler supports extensions that provide array notation for C/C++
  code, as detailed here: http://intel.ly/g454zp}.  Furthermore,
Fortran should be of interest to those studying parallel programming
because of its functional and data parallel constructs and because of
the coarray notation introduced in Fortran 2008. Unlike languages like
C and C++, Fortran has become a truly parallel language with features
added to recent language standards.

@@

%% NEW - CER
History has also shown that an investment in rewriting code does not
guarantee success either, as seen in an effort at LANL to modernize a
legacy Fortran code with the the newer C++ POOMA framework.  This
massive overhaul effort led to a code that was both slower and less
flexible than the original Fortran \cite{basili08hpc}.  Similar
experiences have occurred in the past, notably during the devlopment
of the functional SISAL language in the early 1990s.

### 2010-12-19 opencl/papers/programming-model.tex
edits

\subsubsection*{Regions}

Borrowing ideas from ZPL, we introduce region functions to Fortran.
In Fortran, these functions return a copy of an
existing array or array section.  This is unlike ZPL where regions
are first-class citizens and have special syntax.
These functions imploy a halo (or ghost) cell pattern,
where the size of an array is increased to provide extra array elements
surrounding the interior portion of the array.
%parlab.eecs.berkeley.edu/wiki/_media/patterns/ghostcell.pdf

@@

Unlike OpenMP, we require limited use of compiler directives, although
two are used to enforce the semantics required by the programming
model.  These directives are:

@@

@@

\begin{itemize}
\item There are no loops or index variables to keep track of.  Off by
  one index errors and improper handling of array boundaries are a
  common programming mistake.
\item The written code is closer to the algorithm, easier to
  understand, and is usually substantially shorter.
\item Semantically the intrinsic functions return arrays by value.
  This is usually what the algorithm requires.
\item Because pure and elemental function are free from side effects,
  it is easier for a compiler to schedule the work to be done in
  parallel.
\end{itemize}

@@

Unfortunately, this style of programming has never really caught on because
when Fortran 90 was first introduced, performance was relatively poor and thus
programmers shied away from using array syntax (even now, some are actively
counseling against its usage because of performance issues~\cite{Levesque:SC08}).
Thus the Fortran community was caught in a classic ``chicken-and-egg''
conundrum: (1) programmers didn't use it because it was slow; and (2)
compilers vendors didn't improve it because programmers didn't use it.
A goal of this paper is to demonstrate that parallel programs written in
this style of Fortran are maintainable and can achieve good performance on
accelerator architectures.

@@

Only Fortran procedures are transformed into OpenCL kernels.  The programmer
must explicitly call these kernels from Fortran using the ForOpenCL
library described below.  It is also possible, using ROSE, to modify the calling
site so that the entire program can be transformed, but this functionality is
outside the scope of this paper.  Here we specifically examine transforming
Fortran procedures to OpenCL kernels.  Because OpenCL support is relatively
new to ROSE, some generated code must be modified.  For example, the
{\tt \_\_global} attribute for kernel arguments was added by hand.

@@

Only elemental functions may be called from kernel functions.  These include
fortran functions that have an OpenCL analog and user-defined elemental functions.

A kernel procedure (specified by the {\tt \!\$OFP KERNEL} directive) must be
pure and all array variables must be declared as contiguous.  A kernel
procedure may not call other procedures except for limited intrisic
functions (primarily math), user-defined elemental function, and the
{\tt interior} and {\tt region} functions.

>>>>

@@

Unlike OpenCL, we require limited use of compiler directives, although
two are used to enforce the semantics required by the programming
model.  These directives are:

@@

Borrowing ideas from ZPL, we introduce a concept of a region to
Fortran with a set of functions that allow programmers to work with
subarrays in expressions.  In Fortran, these functions return a copy
of an existing array or array section.  This is unlike ZPL where
regions are analogous to index sets and are used primarily for address
resolution within an array without dictating storage related behavior.
The functions that we propose are similar in that they allow a
programmer to deal with index regions that are meaningful to their
algorithm, and automatically induce a halo (or ghost) cell pattern as
needed in the implementation generated by the compiler, where the size
of an array is increased to provide extra array elements surrounding
the interior portion of the array.  This is critical to reducing the
amount of code that the programmer is forced to write, as the halo cells
are often where boundary errors and off-by-one errors may occur in code
that is manually generated.

@@

\begin{itemize}
\item There are no loops or index variables to keep track of.  Off by
  one index errors and improper handling of array boundaries are a
  common programming mistake.
\item The written code is closer to the algorithm, easier to
  understand, and is usually substantially shorter.
\item Semantically the intrinsic functions return arrays by value.
  This is usually what the algorithm requires.
\item Pure and elemental function are free from side effects, so it is
  easier for a compiler to schedule the work to be done in parallel.
\end{itemize}

@@

Unfortunately, this style of programming has never really caught on
because when Fortran 90 was first introduced, performance of these
features was relatively poor and thus programmers shied away from
using array syntax (even recently, some are actively counseling against its
usage because of performance issues~\cite{Levesque:SC08}).  Thus the
Fortran community was caught in a classic ``chicken-and-egg''
conundrum: (1) programmers didn't use it because it was slow; and (2)
compilers vendors didn't improve it because programmers didn't use it.
A goal of this paper is to demonstrate that parallel programs written
in this style of Fortran are maintainable and can achieve good
performance on accelerator architectures.

@@

Only pure Fortran procedures are transformed into OpenCL kernels.  The
programmer must explicitly call these kernels from Fortran using the
ForOpenCL library described below.  It is also possible, using ROSE,
to modify the calling site so that the entire program can be
transformed, but this functionality is outside the scope of this
paper.  Here we specifically examine transforming Fortran procedures
to OpenCL kernels.  Because OpenCL support is relatively new to ROSE,
some generated code must be modified.  For example, the {\tt
  \_\_global} attribute for kernel arguments was added by hand.

@@

Only elemental functions may be called from kernel functions.  These
include Fortran functions that have an OpenCL analog and user-defined
elemental functions.

A kernel procedure (specified by the {\tt \!\$OFP KERNEL} directive) must be
pure and all array variables must be declared as contiguous.  A kernel
procedure may not call other procedures except for limited intrisic
functions (primarily math), user-defined elemental function, and the
{\tt interior} and {\tt region} functions.  We plan to address
non-contiguous arrays (such as those that result from strided access)
by mapping array strides to gather/scatter-style memory accessors.

### 2010-12-19 opencl/papers/shallow-water.tex
edits

\section{Shallow Water Model}

The numerical code used for this study is from a presentation at the
NM Supercomputing Challenge~\cite{Robey07}.
The algorithm solves the standard 2D Shallow Water equations. This
algorithm is typical of a wide range of modeling equations based on
conservation laws such as compressible fluid dynamics (CFD), elastic
material waves, acoustics, electromagnetic waves and even traffic
flow~\cite{Leveque02}. For the shallow water problem there are
three equations with one based on conservation of mass and the other
two on conservation of momentum.\begin{eqnarray*}
h_{t}+(hu)_{x}+(hv)_{y} & = & 0\quad\mbox{(mass)}\\
(hu)_{t}+(h{u}^{2}+\tfrac{1}{2}gh^{2})_{x}+(huv)_{y} & = & 0\mbox{\quad($x$-momentum)}\\
(hv)_{t}+(huv)_{x}+(h{v}^{2}+\tfrac{1}{2}gh^{2})_{y} & = & 0\mbox{\quad($y$-momentum)}\end{eqnarray*}

@@

\noindent
where h = height of water column (mass), $u$ = x velocity, $v$ =
y velocity, and $g$ = gravity. The height $h$ can be used for mass
because of the simplification of a unit cell size and a uniform water
density. Another simplifying assumption is that the water depth is
small in comparison to length and width and so velocities in the z-direction
can be ignored. A fixed time step is used for simplicity though it
must be less than $dt \leqq dx / (\sqrt{gh}+|u|)$.

>>>>

\section{Shallow Water Model}
\label{sec:shallow-water}

The numerical code used for this work is from a presentation at the
NM Supercomputing Challenge~\cite{Robey07}.
The algorithm solves the standard 2D shallow water equations. This
algorithm is typical of a wide range of modeling equations based on
conservation laws such as compressible fluid dynamics (CFD), elastic
material waves, acoustics, electromagnetic waves and even traffic
flow~\cite{Leveque02}. For the shallow water problem there are
three equations with one based on conservation of mass and the other
two on conservation of momentum.

\begin{eqnarray*}
h_{t}+(hu)_{x}+(hv)_{y} & = & 0\quad\mbox{(mass)}\\
(hu)_{t}+(h{u}^{2}+\tfrac{1}{2}gh^{2})_{x}+(huv)_{y} & = & 0\mbox{\quad($x$-momentum)}\\
(hv)_{t}+(huv)_{x}+(h{v}^{2}+\tfrac{1}{2}gh^{2})_{y} & = & 0\mbox{\quad($y$-momentum)}
\end{eqnarray*}

@@

\noindent
where h = height of water column (mass), $u$ = x velocity, $v$ =
y velocity, and $g$ = gravity. The height $h$ can be used for mass
because of the simplification of a unit cell size and a uniform water
density. Another simplifying assumption is that the water depth is
small in comparison to length and width and so velocities in the z-direction
can be ignored. A fixed time step is used for simplicity though it
must be less than $dt \leqq dx / (\sqrt{gh}+|u|)$ to fulfil the CFL
condition.

### 2010-12-19 opencl/papers/source-to-source.tex
edits

This sections outlines the OpenCL equivalent syntax for portions of the
Fortran shallow-water code described previously.  The notation
uses uppercase for arrays and lowercase for scalar quantities.  Interior
variables are denoted by an i preceeding the array name.  For
example, {\tt iH = interior(H, halo)} is a local copy of interior region of
array {\tt H} (representing height) in the shallow water model.

>>>>

This sections outlines the OpenCL equivalent syntax for portions of
the Fortran shallow-water code described in
Section~\ref{sec:shallow-water}.  The notation uses uppercase for
arrays and lowercase for scalar quantities.  Interior variables are
denoted by an {\tt i} preceeding the array name.  For example, {\tt iH =
  interior(H, halo)} is a local copy of interior region of array {\tt
  H} (representing height) in the shallow water model.
